Informative Autonomous Sampling of Oceanographic Variables Using Joint Excursion Sets

### 2019-10-09 main_new.tex
Updated small fixes on wordings

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop. Myopic sampling, i.e. step-wise selection of
the path (on the graph) which reduces the information criterion the
most, has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or mutual information (e.g. entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumptions of GP models; the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest in \cite{Low2009} and \cite{fossuminformation}.
In this paper, with a focus on mapping the river plume, we reward the
designs that improve the classification of water masses as a means to
characterize the frontal zone and thereby the extent of the river plume.

@@

Data are gathered by an AUV that can measure temperature and salinity
at a subset of locations on the defined grid. This can be done in a
static manner (where the survey design is pre-planned), or in a
sequential way with several rounds of data gathering. For the latter,
a round of data is gathered first, and then assimilated to get an
updated model. Then the updated model is used to design and select the
next round of measurements, and so on (Section
\ref{sec:heuristics}). Below, we describe the model for one round of
data and the Gaussian formula for a single update.

We denote the data at $n_y$ locations by
$\by=(y_{T,1},y_{S,1},\ldots,y_{T,n_y},y_{S,n_y})^t$, representing the
temperature and salinity measurements gathered at the first round of
nodes in the grid. The conditional model for the data, given the true temperature
and salinity, is described by: 
\begin{equation}\label{likelihood}
\by | \bxi \sim N( \bG \bxi, \bR), %\pi(\by | \bxi)= \frac{1}{(2\pi)^{m}|\bR|^{\frac{1}{2}}}e^{-\frac{1}{2}(\by -\bG\bxi)^T\bR^{-1}(\by-\bG\bxi)},
\end{equation}
where $\bG$ is a $2n_y \times 2n$ matrix containing '$1$'s on the
measurement indices of the grid and '$0$' otherwise. The size
$2n_y \times 2n_y$ covariance matrix $\bR$ contains diagonal entries
$r^2_T$ and $r^2_S$ which are measurement error variances of
temperature and salinity observations.

@@

An adaptive AUV survey is split into many rounds (stages). The selection --on where to sample-- is made on the graph and defined by the nearest grid nodes in the domain $\mathcal{M}_g$.

@@

At stage $s$, based on the currently available data
$\mathcal{Y}^{s-1}$, we fit an updated Gaussian model from
Eq. (\ref{gp_upd}) with mean $\bm^{s-1}=E(\bxi|\mathcal{Y}^{s-1})$
and covariance matrix $\bA^{s-1}=\mbox{Var}(\bxi|\mathcal{Y}^{s-1})$.
The next round of measurements $\by^{j,s}$ can be gathered with
design $d^{j,s}$, $j=1,\ldots,J$. The {\it{naive}} strategy then
selects the design according to:
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} |p(\bx_{d^{j,s}};\mathcal{Y}^{s-1})-0.5|, \\
    p(\bx;\mathcal{Y}^{s-1}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S | \mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}
This strategy does not account for the uncertainty in the temperature
or salinity variables, but is instead applicable only if one design (node) has EPs closer to
$0.5$ (see Table \ref{tab:sim_rhoab}, line two). Additionally, this strategy does not
account for spatial correlation. This strategy lacks memory of where
it has been and where the uncertainty has been reduced, and is therefore
susceptible to local minima.

@@

The myopic (greedy) strategy which we present here is optimal if we
imagine taking only one more round of measurements. In this selection
strategy, there is no anticipation of what the subsequent designs might
offer beyond the first round.

Based on the currently available data $\mathcal{Y}^{s-1}$, we
fit an updated GP model.  The next round of measurements
$\by^{j,s}$, can be gathered with design $d^{j,s}$, $j=1,\ldots,J$. The
selected design is then:
\begin{eqnarray}\label{critSEQ}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} \left\{ V_{m,\mbox{upd,j}} \right\},  \\
V_{m,\mbox{upd}} & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by^{j,s}|\mathcal{Y}^{s-1}} \left( p(\bx;\mathcal{Y}^{j,s})\left( 1-p(\bx;\mathcal{Y}^{j,s})\right) \right) \Delta, \nonumber \\
    p(\bx;\mathcal{Y}^{j,s}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by^{j,s},\mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}

Note that this strategy gives a sequential conditional version of the
formula in Eq. (\ref{sur}). Now $\mathcal{Y}^{s-1}$ is available, and
the expectation is with respect to the conditional density
$\pi(\by^{j,s}|\mathcal{Y}^{s-1})$. A similar closed-form calculation
for expected IBV is hence applicable in Eq. (\ref{critSEQ}) using the
updated GP model from step $s-1$. Once the data is collected for the
best design, the GP model is updated again. The mean $\bm^{s}$ and
covariance matrix $\bA^{s}$ are used to compute the next design at
stage $s+1$, and so on.

@@

We now extend the myopic strategy to a look-ahead strategy, which is
optimal when one can gather only two more rounds of measurements. In
addition to the next round of measurements $\by^{j,s}$, this
look-ahead strategy, anticipates the subsequent design $j_2$ with
data $\by^{j_2,s+1}$ when choosing the current design $d^{j,s}$.  The
selected design is:
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} \left\{ U_{la,\mbox{upd,j}} \right\},  \\
    U_{la,\mbox{upd},j} & = &  E_{\by^{j,s}|\mathcal{Y}^{s-1}} \left\{ \mbox{argmin}_{j_2 \in \{1,\ldots,{J}_2\}} \left[ V_{la,\mbox{upd},j_2} \right] \right\}, \nonumber \\
V_{la,\mbox{upd},j_2} & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by^{j_2,s+1}|\mathcal{Y}^{j,s}} \left\{ p(\bx;\mathcal{Y}^{j,s+1})\left( 1-p(\bx;\mathcal{Y}^{j,s+1})\right) \right\} \Delta, \nonumber \\
    p(\bx;\mathcal{Y}^{j,s+1}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by^{j_2,s+1},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by^{j,s},\mathcal{Y}^{s-1}\}$ and
$\mathcal{Y}^{j,s+1}=\{\by^{j_2,s+1},\mathcal{Y}^{j,s}\}$ represent
the sets of data variables, and the expectations are calculated with respect to
the conditional densities $\pi(\by^{j,s}|\mathcal{Y}^{s-1})$ and
$\pi(\by^{j_2,s+1}|\by^{j,s},\mathcal{Y}^{s-1})$.

@@

\begin{figure}[h!]
  \centering
  % \subfigure[Excursion set variance $E_{\by}(p[1-p])$.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \subfigure[IBV.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \hfill
  \subfigure[RMSE between estimated field and truth.]{\label{fig:avg_rmse}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_RMSE.pdf}}
  \hfill 
  \subfigure[Explained variance $\bR^{2}$.]{\label{fig:avg_r2}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_R2.pdf}}
  \hfill 
  \subfigure[Time used to do inference.]{\label{fig:avg_time}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_Time.pdf}}
\caption{Simulation results from 100 replicate simulations for 10
  sampling choices/steps on the grid.} 
\label{fig:sim_results}
\end{figure}

@@

\begin{figure}[!b]
  \centering
  \subfigure[Look-ahead]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_look-ahead.pdf}\label{fig:avg_look-ahead}}
  \hfill
  \subfigure[Myopic]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_myopic.pdf}\label{fig:avg_myopic}}
  \hfill
  \subfigure[Naive]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_naive.pdf}\label{fig:route_naive}}
  \hfill
  \subfigure[Static]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_static.pdf}\label{fig:route_static}}
  \hfill
  \caption{The overlaid route choices (strategies), superimposed on
    the 100 replicate surveys with 10 sampling choices/steps.}
\label{fig:route_choices}
\end{figure}

@@

The tables below explore the sensitivity of the result to different
prior correlation values between temperature measured onboard the AUV,
temperature and salinity, prior standard deviations, and spatial
correlations (see table rows). We only compare results at the end of
each survey. The expected ES variances are shown in Table
\ref{tab:sim_res_ev}, results of RMSE are in Table
\ref{tab:sim_res_rmse}, and explained variance is in Table
\ref{tab:sim_res_r2}. In all runs, both the \textit{myopic} and
\textit{look-ahead} strategies perform the best in terms of expected
variance in the ES. Surprisingly, the \textit{myopic} strategy
performs slightly better than the \textit{look-ahead}. Part of the
reason for this discrepancy can be traced to the act of pruning some
of the paths for the \textit{look-ahead} strategy; however, even with
less pruning, the \textit{look-ahead} does not outperform the
\textit{myopic} approach. The \textit{static\_north} design is clearly
better than the other two static designs, the reason being that the
others do not focus on the areas where the ES is really uncertain
(centre locations are the most uncertain). By measuring temperature
alone, the expected variances in the ES are only slightly higher than
the combined temperature-salinity estimate. This occurs because the
correlation of $0.6$ means that temperature observations provide
information about salinity as well. For the other predictive measures,
\textit{myopic} and \textit{look-ahead} strategies do well, clearly
beating the \textit{naive} strategy.

@@

While more effort could be spent on approximating the look-ahead steps
in the adaptive designs, it is perhaps more interesting to explore the
additional flexibility that can be gained by having two or more AUVs
co-temporally exploring a spatial or spatio-temporal domain together
\citep{ferreira2019advancing}. Such an approach would enable
concurrent sampling in different parts of the space, or opportunities
to move in parallel to best capture the excursion set.

>>>>

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop. Myopic sampling, i.e. stage-wise selection of
the path (on the graph) which reduces the information criterion the
most, has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or mutual information (e.g. entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumptions of GP models; the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest in \cite{Low2009} and \cite{fossuminformation}.
In this paper, with a focus on mapping the river plume, we reward the
designs that improve the classification of water masses as a means to
characterize the frontal zone and thereby the extent of the river plume.

@@

Data are gathered by an AUV that can measure temperature and salinity
at a subset of locations on the defined grid. This can be done in a
static manner (where the survey design is pre-planned), or in a
sequential way with several stages of data gathering. For the latter,
a stage of data is gathered first, and then assimilated to get an
updated model. Then the updated model is used to design and select the
next stage of measurements, and so on (Section
\ref{sec:heuristics}). Below, we describe the model for one stage of
data and the Gaussian formula for a single update.

We denote the data at $n_y$ locations by
$\by=(y_{T,1},y_{S,1},\ldots,y_{T,n_y},y_{S,n_y})^t$, representing the
temperature and salinity measurements gathered at the first stage of
nodes in the grid. The conditional model for the data, given the true temperature
and salinity, is described by: 
\begin{equation}\label{likelihood}
\by | \bxi \sim N( \bG \bxi, \bR), %\pi(\by | \bxi)= \frac{1}{(2\pi)^{m}|\bR|^{\frac{1}{2}}}e^{-\frac{1}{2}(\by -\bG\bxi)^T\bR^{-1}(\by-\bG\bxi)},
\end{equation}
where $\bG$ is a $2n_y \times 2n$ matrix containing '$1$'s on the
measurement indices of the grid and '$0$' otherwise. The size
$2n_y \times 2n_y$ covariance matrix $\bR$ contains diagonal entries
$r^2_T$ and $r^2_S$ which are measurement error variances of
temperature and salinity observations.

@@

An adaptive AUV survey is split into many stages. The selection --on where to sample-- is made on the graph and defined by the nearest grid nodes in the domain $\mathcal{M}_g$.

@@

At stage $s$, based on the currently available data
$\mathcal{Y}^{s-1}$, we fit an updated Gaussian model from
Eq. (\ref{gp_upd}) with mean $\bm^{s-1}=E(\bxi|\mathcal{Y}^{s-1})$
and covariance matrix $\bA^{s-1}=\mbox{Var}(\bxi|\mathcal{Y}^{s-1})$.
The next stage of measurements $\by^{j,s}$ can be gathered with
design $d^{j,s}$, $j=1,\ldots,J$. The {\it{naive}} strategy then
selects the design according to:
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} |p(\bx_{d^{j,s}};\mathcal{Y}^{s-1})-0.5|, \\
    p(\bx;\mathcal{Y}^{s-1}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S | \mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}
This strategy does not account for the uncertainty in the temperature
or salinity variables, but is instead applicable only if one design (node) has EPs closer to
$0.5$ (see Table \ref{tab:sim_rhoab}, line two). Additionally, this strategy does not
account for spatial correlation. This strategy lacks memory of where
it has been and where the uncertainty has been reduced, and is therefore
susceptible to local minima.

@@

The myopic (greedy) strategy which we present here is optimal if we
imagine taking only one more stage of measurements. In this selection
strategy, there is no anticipation of what the subsequent designs might
offer beyond the first stage.

Based on the currently available data $\mathcal{Y}^{s-1}$, we
fit an updated GP model.  The next stage of measurements
$\by^{j,s}$, can be gathered with design $d^{j,s}$, $j=1,\ldots,J$. The
selected design is then:
\begin{eqnarray}\label{critSEQ}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} \left\{ V_{m,\mbox{upd,j}} \right\},  \\
V_{m,\mbox{upd}} & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by^{j,s}|\mathcal{Y}^{s-1}} \left( p(\bx;\mathcal{Y}^{j,s})\left( 1-p(\bx;\mathcal{Y}^{j,s})\right) \right) \Delta, \nonumber \\
    p(\bx;\mathcal{Y}^{j,s}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by^{j,s},\mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}

Note that this strategy gives a sequential conditional version of the
formula in Eq. (\ref{sur}). Now $\mathcal{Y}^{s-1}$ is available, and
the expectation is with respect to the conditional density
$\pi(\by^{j,s}|\mathcal{Y}^{s-1})$. A similar closed-form calculation
for expected IBV is hence applicable in Eq. (\ref{critSEQ}) using the
updated GP model from stage $s-1$. Once the data is collected for the
best design, the GP model is updated again. The mean $\bm^{s}$ and
covariance matrix $\bA^{s}$ are used to compute the next design at
stage $s+1$, and so on.

@@

We now extend the myopic strategy to a look-ahead strategy, which is
optimal when one can gather only two more stages of measurements. In
addition to the next stage of measurements $\by^{j,s}$, this
look-ahead strategy, anticipates the subsequent design $j_2$ with
data $\by^{j_2,s+1}$ when choosing the current design $d^{j,s}$.  The
selected design is:
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} \left\{ U_{la,\mbox{upd,j}} \right\},  \\
    U_{la,\mbox{upd},j} & = &  E_{\by^{j,s}|\mathcal{Y}^{s-1}} \left\{ \mbox{argmin}_{j_2 \in \{1,\ldots,{J}_2\}} \left[ V_{la,\mbox{upd},j_2} \right] \right\}, \nonumber \\
V_{la,\mbox{upd},j_2} & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by^{j_2,s+1}|\mathcal{Y}^{j,s}} \left\{ p(\bx;\mathcal{Y}^{j,s+1})\left( 1-p(\bx;\mathcal{Y}^{j,s+1})\right) \right\} \Delta, \nonumber \\
    p(\bx;\mathcal{Y}^{j,s+1}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by^{j_2,s+1},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by^{j,s},\mathcal{Y}^{s-1}\}$ and
$\mathcal{Y}^{j,s+1}=\{\by^{j_2,s+1},\mathcal{Y}^{j,s}\}$ represent
the sets of data variables, and the expectations are calculated with respect to
the conditional densities $\pi(\by^{j,s}|\mathcal{Y}^{s-1})$ and
$\pi(\by^{j_2,s+1}|\by^{j,s},\mathcal{Y}^{s-1})$.

@@

\begin{figure}[h!]
  \centering
  % \subfigure[Excursion set variance $E_{\by}(p[1-p])$.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \subfigure[IBV.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \hfill
  \subfigure[RMSE between estimated field and truth.]{\label{fig:avg_rmse}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_RMSE.pdf}}
  \hfill 
  \subfigure[Explained variance $\bR^{2}$.]{\label{fig:avg_r2}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_R2.pdf}}
  \hfill 
  \subfigure[Time used to do inference.]{\label{fig:avg_time}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_Time.pdf}}
\caption{Simulation results from 100 replicate simulations for 10
  sampling choices/stages on the grid.} 
\label{fig:sim_results}
\end{figure}

@@

\begin{figure}[!b]
  \centering
  \subfigure[Look-ahead]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_look-ahead.pdf}\label{fig:avg_look-ahead}}
  \hfill
  \subfigure[Myopic]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_myopic.pdf}\label{fig:avg_myopic}}
  \hfill
  \subfigure[Naive]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_naive.pdf}\label{fig:route_naive}}
  \hfill
  \subfigure[Static]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_static.pdf}\label{fig:route_static}}
  \hfill
  \caption{The overlaid route choices (strategies), superimposed on
    the 100 replicate surveys with 10 sampling choices/stages.}
\label{fig:route_choices}
\end{figure}

@@

The tables below explore the sensitivity of the result to different
prior correlation values between temperature measured onboard the AUV,
temperature and salinity, prior standard deviations, and spatial
correlations (see table rows). We only compare results at the end of
each survey. The expected ES variances are shown in Table
\ref{tab:sim_res_ev}, results of RMSE are in Table
\ref{tab:sim_res_rmse}, and explained variance is in Table
\ref{tab:sim_res_r2}. In all runs, both the \textit{myopic} and
\textit{look-ahead} strategies perform the best in terms of expected
variance in the ES. The \textit{myopic} strategy
performs slightly better than the \textit{look-ahead}. Part of the
reason for this discrepancy can be traced to the act of pruning some
of the paths for the \textit{look-ahead} strategy. The \textit{static\_north} design is clearly
better than the other two static designs, the reason being that the
others do not focus on the areas where the ES is really uncertain
(centre locations are the most uncertain). By measuring temperature
alone, the expected variances in the ES are only slightly higher than
the combined temperature-salinity estimate. This occurs because the
correlation of $0.6$ means that temperature observations provide
information about salinity as well. For the other predictive measures,
\textit{myopic} and \textit{look-ahead} strategies do well, clearly
beating the \textit{naive} strategy.

@@

While more effort could be spent on approximating the look-ahead stages
in the adaptive designs, it is perhaps more interesting to explore the
additional flexibility that can be gained by having two or more AUVs
co-temporally exploring a spatial or spatio-temporal domain together
\citep{ferreira2019advancing}. Such an approach would enable
concurrent sampling in different parts of the space, or opportunities
to move in parallel to best capture the excursion set.
